---
title: "ICU-VTE Score validation"
author: "SCS Rivrud, RJ Eck, ERH Heijkoop"
output:
  html_document:
    code_folding: hide
---

**Load packages**
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(kableExtra)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(DBI)
library(RSQLite)
library(gtsummary)
library(reshape2)
library(dbplyr)
library(stringr)
library(readxl)
library(writexl)
library(stringi)
library(rms)
library(gmish)
library(glmnet)
library(pec)
library(Hmisc)
library(qreport)
library(DescTools)
library(e1071)
library(ggpubr)
library(bestNormalize)
library(GGally)
library(MASS)
library(pROC)
library(rsample)
library(riskRegression)
library(here)
library(tidyverse)
library(RSQLite)
library(survival)
library(survminer)
library(dplyr)
library(gridExtra)
library(grid)
library(adjustedCurves)
library(dcurves)
library(mice)
```

**Load datasets and perform multiple imputation**
```{r eval=FALSE}
mimic <- readRDS("analysis subsets/icu_vte_mimic.rds")%>%
  mutate(platelets_scaled = platelets/100) #scale platelets, according to model by viarasilpa

number_df <- 10 #specify nr imputed dfs

#perform imputation
mimic_imp <- mice(mimic, m=number_df, maxit=30, method = "cart", printFlag = TRUE) 
```

**Boostrap function for internal validation**
```{r eval=FALSE}
#define bootstrap function
bootstrap_validation_logreg_external <- function(data, n_bootstrap) {
  
  #make storage for apparent values (original model tested on bootstrap dataset)
  auc_values_apparent <- numeric(n_bootstrap)
  brier_score_apparent <- numeric(n_bootstrap) 
  emax_apparent <- numeric(n_bootstrap)
  e50_apparent <- numeric(n_bootstrap)
  e90_apparent <- numeric(n_bootstrap)
  r2_apparent <- numeric(n_bootstrap)
  ici_apparent <- numeric(n_bootstrap)

  for(i in 1:n_bootstrap) {
    #resampling indices
    bootstrap_data <- sample_n(data, nrow(data), replace = TRUE)

    #predict on the bootstrap sample
    bootstrap_data <- bootstrap_data %>%
      mutate(#based on model in the supplemntary material of viarasilpa et al
         odds_vte = exp(-7.1142 + 2.0308*cvl_num + 0.8928*mech_vent_num + 1.4384*previousvte_num + 0.13103*platelets_scaled + 0.0412*total_immobile_days + 0.8020*hb_cutoff + (-0.8078*ffp_num)),
         
         prob_vte = odds_vte / (1 + odds_vte) #calculate probabilities based on odds
         )
    
    #store apparent values
    auc_result_apparent <- roc(bootstrap_data$status_dich, bootstrap_data$prob_vte, ci = TRUE, direction = "<")
    apparent_cal <- val.prob(bootstrap_data$prob_vte, bootstrap_data$status_dich)
    auc_values_apparent[i] <- auc_result_apparent$auc
    brier_score_apparent[i] <- mean((bootstrap_data$prob_vte - as.numeric(bootstrap_data$status_dich))^2)
    emax_apparent[i] <- apparent_cal[[14]]
    e50_apparent[i] <- unname(quantile(abs(bootstrap_data$prob_vte - approx(lowess(bootstrap_data$prob_vte, bootstrap_data$status_dich, iter=0), xout=bootstrap_data$prob_vte, ties=mean)$y), 0.5))
    e90_apparent[i] <- apparent_cal[[15]] 
    r2_apparent[i] <- apparent_cal[[3]] 
    ici_apparent[i] <- apparent_cal[[16]]
    
  }
  
  #pooling the results directly in the function
  auc_df <- data.frame(
    auc_apparent = auc_values_apparent,
    brier_score_apparent = brier_score_apparent,
    emax_apparent = emax_apparent,
    e50_apparent = e50_apparent,
    e90_apparent = e90_apparent,
    r2_apparent = r2_apparent,
    ici_apparent = ici_apparent
    
  )
  
  #summary of results 
  auc_df_summary <- auc_df %>%
    summarise(
     m_auc_apparent = mean(auc_apparent),
    se_auc_apparent = sd(auc_apparent, na.rm = TRUE)/sqrt(n_bootstrap),
    
    m_brier_score_apparent = mean(brier_score_apparent, na.rm = TRUE),
    se_brier_score_apparent = sd(brier_score_apparent, na.rm = TRUE)/sqrt(n_bootstrap),
    
    m_e50_apparent = mean(e50_apparent, na.rm = TRUE),
    se_e50_apparent = sd(e50_apparent)/sqrt(n_bootstrap),

    m_emax_apparent = mean(emax_apparent, na.rm = TRUE),
    se_emax_apparent = sd(emax_apparent, na.rm = TRUE)/sqrt(n_bootstrap),
 
    m_e90_apparent = mean(e90_apparent, na.rm = TRUE),
    se_e90_apparent = sd(e90_apparent)/sqrt(n_bootstrap),

    m_r2_apparent = mean(r2_apparent, na.rm = TRUE),
    se_r2_apparent = sd(r2_apparent)/sqrt(n_bootstrap),
    
    m_ici_apparent = mean(ici_apparent, na.rm = TRUE),
    se_ici_apparent = sd(ici_apparent)/sqrt(n_bootstrap)

    )
  
  return(auc_df_summary)
}
```

**Running the boostrap validation and making summary tables**
```{r eval=FALSE}

n_bootstrap <- 200
results_list <- list()

  #loop through each imputed dataset
for (i in 1:number_df) {
  #complete the dataset
  data_i <- complete(mimic_imp, action = i)%>%
  mutate(platelets_scaled = platelets/100)
    #apply the bootstrap validation function
  results_list[[i]] <- bootstrap_validation_logreg_external(data_i, n_bootstrap)
}

#combine all results into a single data frame
results_df <- bind_rows(results_list)

#pool the results according to rubins rules
final_results_external_validation <- results_df %>%
 summarise(
    auc_apparent = mean(m_auc_apparent),
    se_auc_apparent = sqrt(sd(m_auc_apparent, na.rm = TRUE)^2 + mean(se_auc_apparent^2) + (sd(m_auc_apparent, na.rm = TRUE)^2)/number_df),
    
    brier_score_apparent = mean(m_brier_score_apparent, na.rm = TRUE),
    se_brier_score_apparent = sqrt(sd(m_brier_score_apparent, na.rm = TRUE)^2 + mean(se_brier_score_apparent^2) + (sd(m_brier_score_apparent, na.rm = TRUE)^2)/number_df),
    
    e50_apparent = mean(m_e50_apparent, na.rm = TRUE),
    se_e50_apparent = sqrt(sd(m_e50_apparent, na.rm = TRUE)^2 + mean(se_e50_apparent^2) + (sd(m_e50_apparent, na.rm = TRUE)^2)/number_df),
    
    emax_apparent = mean(m_emax_apparent, na.rm = TRUE),
    se_emax_apparent = sqrt(sd(m_emax_apparent, na.rm = TRUE)^2 + mean(se_emax_apparent^2) + (sd(m_emax_apparent, na.rm = TRUE)^2)/number_df),
    
    e90_apparent = mean(m_e90_apparent, na.rm = TRUE),
    se_e90_apparent = sqrt(sd(m_e90_apparent, na.rm = TRUE)^2 + mean(se_e90_apparent^2) + (sd(m_e90_apparent, na.rm = TRUE)^2)/number_df),
    
    r2_apparent = mean(m_r2_apparent, na.rm = TRUE),
    se_r2_apparent = sqrt(sd(m_r2_apparent, na.rm = TRUE)^2 + mean(se_r2_apparent^2) + (sd(m_r2_apparent, na.rm = TRUE)^2)/number_df),
    
    ici_apparent = mean(m_ici_apparent, na.rm = TRUE),
    se_ici_apparent = sqrt(sd(m_ici_apparent, na.rm = TRUE)^2 + mean(se_ici_apparent^2) + (sd(m_ici_apparent, na.rm = TRUE)^2)/number_df)
  )

#extract metrics
df_metrics <- pivot_longer(final_results_external_validation, 
                        cols = -starts_with("se_"), 
                        names_to = "metric", 
                        values_to = "value")

#extract the se columns into a separate data frame
df_se <- pivot_longer(final_results_external_validation, 
                      cols = starts_with("se_"), 
                      names_to = "metric_se", 
                      values_to = "se")

#remove 'se_' prefix to match metric names so we can left join them
df_se$metric_se <- sub("se_", "", df_se$metric_se)

#merge the value and se columns by matching metric names
all_metrics_df <- left_join(df_metrics, df_se, by = c("metric" = "metric_se"))

#rename the columns appropriately
all_metrics_df <- all_metrics_df %>% dplyr::select(metric, value, se)

#make printable df
printable_external_validation <- all_metrics_df %>%
  filter(metric != "r2_apparent") %>%
  mutate(
    value = signif(value, digits = 3),
    lower_ci = signif(value - se*1.96, digits = 3),
    upper_ci= signif(value + se*1.96, digits = 3),
     `Value (95% CI)` = paste(value, paste("(", lower_ci, "-",upper_ci, ")", sep = "")))%>%
  dplyr::select(Metric = metric, `Apparent value (95% CI)` = `Value (95% CI)`)%>%
   mutate(Metric = case_when(
    Metric == "auc_apparent" ~ "C-statistic",
    Metric == "brier_score_apparent" ~ "Brier score",
    Metric == "emax_apparent" ~ "Emax",
    Metric == "e50_apparent" ~ "E50",
    Metric == "e90_apparent" ~ "E90",
    Metric == "ici_apparent" ~ "ICI"
  ))%>%
  kbl("html", escape = FALSE) %>%
  kable_classic(full_width = F, html_font = "Times")
```

**Making calibration + density plot for external validation**
```{r eval=FALSE}
#calculating the mean predicted risk per subject
mimic_prob <- complete(mimic_imp, "long") %>%
      mutate(#based on model in supplementary material of viarasilpa et al
        platelets_scaled = platelets/100,
         odds_vte = exp(-7.1142 + 2.0308*cvl_num + 0.8928*mech_vent_num + 1.4384*previousvte_num + 0.13103*platelets_scaled + 0.0412*total_immobile_days + 0.8020*hb_cutoff + (-0.8078*ffp_num)),
         
         prob_vte = odds_vte / (1 + odds_vte)
         ) %>%
  group_by(.id) %>%
  summarise(prob_vte = mean(prob_vte),
            status_dich = max(status_dich)) %>%
  ungroup()

percentile_95 <- quantile(mimic_prob$prob_vte, probs = 0.95)

external_validation_1 <- ggplot(mimic_prob, aes(x = prob_vte, y = status_dich)) +
  geom_smooth(color = "darkblue", method = "loess", se = FALSE, size = 0.5, span = 0.75) + #LOESS smooth line
  stat_smooth(color = "darkred", method="glm", family="binomial", se=F, size = 0.5) +
  geom_vline(xintercept = percentile_95, linetype = "dashed", color = "gray", size = 0.5) + #95th percentile line
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.05)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.05)) +
  geom_abline(size = 4, alpha = 0.5, color = "gray") + #adds a line with a slope of 1 and intercept of 0 by default
  xlab("") +
  ylab("Observed Probability") +
  theme_minimal() +
  ggtitle(expression(bold("Figure x.") ~ "Calibration plot of external validation of the ICU-VTE score")) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  theme(plot.title = element_text(hjust = 0.5, family = "Times New Roman", face = "bold"),
        text = element_text(family = "Times New Roman"),
        panel.background = element_rect(fill = "white", colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        axis.title.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        plot.margin = unit(c(1,1,0,1), "lines"))

external_validation_2 <- ggplot(mimic_prob, aes(x = prob_vte)) +
  geom_histogram(color = "black", fill = NA, binwidth = 0.001) + 
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.05)) +
  xlab("Predicted Probability") +
  ylab("") +
  theme_minimal() +
  theme(text = element_text(family = "Times New Roman"),
        panel.background = element_rect(fill = "white", colour = NA), #remove border color
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_text(colour = "white"),
        panel.border = element_blank()) + #add line for x-axis border
  geom_segment(aes(x = 0, y = 0, xend = 1, yend = 0), color = "black", size = 0.05) #add manual border for x-axis from x=0 to x=0.35


legend_plot <- ggplot(data.frame(x = c(0, 1), y = c(0, 1)), aes(x, y)) +
  geom_line(aes(color = "Logistic calibration"), size = 0.5) +
  geom_line(aes(color = "Locally estimated scatterplot smoothing"), size = 0.5) +
  geom_line(aes(color = "95% Percentile of Predicted Risks"), size = 0.5, linetype = "dashed") +
  geom_line(size = 10, show.legend = FALSE, color = "white", fill = "white") + #do not show in legend
  scale_color_manual("", 
                     values = c("Logistic calibration" = "darkred", 
                                "Locally estimated scatterplot smoothing" = "darkblue",
                                "95% Percentile of Predicted Risks" = "gray"
                     )) +
  theme(legend.position = "top", 
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        axis.title.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(), 
        axis.ticks.y = element_blank(),
        axis.text.x= element_blank(),
        axis.text.y = element_blank(),
        panel.border = element_blank(),
        legend.text = element_text(size = 10, family = "Times New Roman"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_blank())



#combining the adjusted plots
external_validation_calibration <- arrangeGrob(external_validation_1, external_validation_2, legend_plot, 
                                               heights = unit(c(3, 1, 0.6), "null"), ncol = 1)
```

**Making calibration + density plot for external validation zoomed in on the 99.9th percentile**
```{r eval=FALSE}
#get 99.9th percentile
cutoff_upper <- quantile(mimic_prob$prob_vte, probs = 0.999)

external_validation_1 <- ggplot(mimic_prob, aes(x = prob_vte, y = status_dich)) +
  geom_smooth(color = "darkblue", method = "loess", se = FALSE, size = 0.5, span = 0.75) + #LOESS smooth line
  stat_smooth(color = "darkred", method="glm", family="binomial", se=F, size = 0.5) +
  geom_vline(xintercept = percentile_95, linetype = "dashed", color = "gray", size = 0.5) + #95th percentile line
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.05)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.05)) +
  geom_abline(size = 4, alpha = 0.5, color = "gray") + #adds a line with a slope of 1 and intercept of 0 by default
  xlab("") +
  ylab("Observed Probability") +
  theme_minimal() +
  ggtitle(expression(bold("Figure x.") ~ "Calibration plot of external validation of the ICU-VTE score")) +
  coord_cartesian(xlim = c(0, 0.4), ylim = c(0, 0.4)) +
  theme(plot.title = element_text(hjust = 0.5, family = "Times New Roman", face = "bold"),
        text = element_text(family = "Times New Roman"),
        panel.background = element_rect(fill = "white", colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        axis.title.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        plot.margin = unit(c(1,1,0,1), "lines"))

external_validation_2 <- ggplot(mimic_prob, aes(x = prob_vte)) +
  geom_histogram(color = "black", fill = NA, binwidth = 0.001) + 
  scale_x_continuous(limits = c(0, 0.4), breaks = seq(0, 0.4, by = 0.05)) +
  xlab("Predicted Probability") +
  ylab("") +
  theme_minimal() +
  theme(text = element_text(family = "Times New Roman"),
        panel.background = element_rect(fill = "white", colour = NA), #remove border color
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_text(colour = "white"),
        panel.border = element_blank()) + #add line for x-axis border
  geom_segment(aes(x = 0, y = 0, xend = 0.4, yend = 0), color = "black", size = 0.05) #add manual border for x-axis from x=0 to x=0.35


legend_plot <- ggplot(data.frame(x = c(0, 1), y = c(0, 1)), aes(x, y)) +
  geom_line(aes(color = "Logistic calibration"), size = 0.5) +
  geom_line(aes(color = "Locally estimated scatterplot smoothing"), size = 0.5) +
  geom_line(aes(color = "95% Percentile of Predicted Risks"), size = 0.5, linetype = "dashed") +
  geom_line(size = 10, show.legend = FALSE, color = "white", fill = "white") + #do not show in legend
  scale_color_manual("", 
                     values = c("Logistic calibration" = "darkred", 
                                "Locally estimated scatterplot smoothing" = "darkblue",
                                "95% Percentile of Predicted Risks" = "gray"
                     )) +
  theme(legend.position = "top", 
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        axis.title.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(), 
        axis.ticks.y = element_blank(),
        axis.text.x= element_blank(),
        axis.text.y = element_blank(),
        panel.border = element_blank(),
        legend.text = element_text(size = 10, family = "Times New Roman"),
        legend.background = element_rect(fill = "white"),
        legend.key = element_blank())



#combining the adjusted plots
external_validation_calibration_zoom <- arrangeGrob(external_validation_1, external_validation_2, legend_plot, 
                                               heights = unit(c(3, 1, 0.6), "null"), ncol = 1)
```

**Save R environment**
```{r eval=FALSE}
save.image(file ='icu_vte_validation_environment.RData')
```

**Load R environment**
```{r}
load('icu_vte_validation_environment.RData')
```

**External validation stats**
```{r}
printable_external_validation
```

**print external validation calibration plot**
```{r fig.height=8, fig.width=8}
grid.newpage()
grid.draw(external_validation_calibration)
```

**print external validation calibration plot zoomed**
```{r fig.height=8, fig.width=8}
grid.newpage()
grid.draw(external_validation_calibration_zoom)
```

**print external validation calibration plot zoomed**
```{r fig.height=8, fig.width=8}
grid.newpage()
grid.draw(external_validation_calibration_zoom_show)
```

